---
title: "Class Project"
subtitle: "Final Report"
author: Branson T Fausey
date: Sunday, 8th December, 2024
output: html_notebook
---

```{r setup, include=FALSE}
# cleanup & stock the library shelves
rm(list = ls())
library(magrittr)
library(mosaic)
library(openintro)
library(tidyverse)
library(tidyr)
library(manipulate)
library(esquisse)
library(ggplot2)
library(ggthemes)
library(dplyr)
library(printr)
library(readr)
library(devtools)
library(lubridate)
library(leaflet)
library(foreign)
library(data.table)
library(base)
library(rvest)
library(tidytuesdayR)
library(usmap)
library(readxl)
set.seed(184)
```

___

# Research Question:

*How has Pennsylvania's deer population changed over time, and how does it compare to other states' populations? How about compared to other popular game animals in Pennsylvania?*

* For clarification, my main interest is the white-tailed deer species, but my analyses and descriptions will include other species of deer, mainly species native to the United States such as white-tailed and mule deer.

___

## Data Sources:

* [Deer Friendly](https://www.deerfriendly.com/)

* [Pennsylvania Deer Harvest 1915 to 2020](https://docs.google.com/spreadsheets/d/1C8gSJWomCEdPSvHPHm7UVd_oGQU6gmsm/edit?gid=1959951972#gid=1959951972)

* [US Deer Population Long-term Chart](https://docs.google.com/spreadsheets/d/1nuAUVkXsp1HkhIlPsiNFG_BzyO5HOYdS01jqrYXX-FY/edit?gid=1580702380#gid=1580702380)

* [Texas Deer Population 2000 to 2021](https://docs.google.com/spreadsheets/d/1GLy-Knmp_IzJALFd01E2bJCG8h0Ow3I4S9nCl7PDnbE/edit?gid=0#gid=0)

* [Nevada Mule Deer Population](https://docs.google.com/spreadsheets/d/1ZIL0-G_3x3nTbz_pz135ttUocQLy2OuIecnBJgdgpIM/edit?gid=0#gid=0)

___

Texas is reported to host the largest deer population in the country, while Nevada is reported to have the least amount of deer for a state in the contiguous US.

The PA and US data sources are cluttered with images and descriptions that follow no sense of data organization, so without changing any values, I simply remove the unnecessary  bits in the Excel files to import.

The TX and NV data sets can be read-in okay without adjustments.

All data is collected by government entities and compiled by Deer Friendly.

___

```{r}
# data import
PAdeerharvest <- read_excel("PAdeerharvest.xlsx")
USdeerpop <- read_excel("USdeerpop.xlsx")
TXdeerpop <- read_excel("TXdeerpop.xlsx")
NVdeerpop <- read_excel("NVdeerpop.xlsx")
```

```{r}
# inspecting data intake
PAdeerharvest %>%head()
USdeerpop %>% head()
TXdeerpop %>% head()
NVdeerpop %>% head()
```

This data is fairly straightforward, though it looks like there are some typography differences in the PADeerharvest data's 'year' variable that we can remedy. Some of the other variables that should be numeric are instead set as character values, so that will have to be fixed as well, should we decide to do any plotting of the data or calculations with it.

Further inspection of the source webpage for the USdeerpop dataset tells us that the values in this dataset are in millions.

It is also clear from my research that Pennsylvania does not publish data on estimated deer population sizes like other states do, but rather publish the harvest data from yearly hunting seasons, using this as an indicator for the deer population. This is unfortunate, but the TXdeerpop dataset features a 'Hunt' variable, so this can be directly compared with data from the PAdeerharvest data.

```{r}

```



